---
title: "Final Project - Part1"
author: "Dan Nguyen"
date: "2026-03-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# 1 - Read Data and EDA 

```{r}
# Read Data
bike_raw <- read.csv("bike_sharing_daily.csv")

# Re-assign values of variables
bike_raw <- bike_raw %>%
  mutate(
    season     = factor(season,     levels = 1:4,  
                        labels = c("Spring", "Summer", "Fall", "Winter")),
    mnth       = factor(mnth,       levels = 1:12),
    holiday    = factor(holiday,    levels = c(0,1), labels = c("No", "Yes")),
    weekday    = factor(weekday,    levels = 0:6,  
                        labels = c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")),
    workingday = factor(workingday, levels = c(0,1), labels = c("No", "Yes")),
    weathersit = factor(weathersit, levels = 1:4,  
                        labels = c("Clear", "Mist/Cloudy", "Light Rain/Snow", "Heavy Rain/Snow"))
  )

# Filter for 2011
bike_2011 <- bike_raw %>% filter(yr == 0)
```

## Categorical Variables  
Season, month, holiday, weekday, and workingday are standard characteristics within a year. There is no reason to expect significant variation in their distributions. Thus, we will not print out summary tables and plots for these variables.  

### Weather Situation  
```{r}
table(bike_2011$weathersit)
prop.table(table(bike_2011$weathersit))
```
```{r}
bike_2011 %>%
  count(weathersit) %>%
  ggplot(aes(x = weathersit, y = n)) +
  geom_col(width = 0.6, color = "white") +
  geom_text(aes(label = n), vjust = -0.4, fontface = "bold") +
  labs(title = "Weather Situation Distribution (2011)",
       x = "Weather Situation", y = "# Days") +
  theme_minimal()
```
## Continuous Variables  
```{r}
cont_vars <- bike_2011 %>% select(temp, atemp, hum, windspeed, casual, registered, cnt)

# Numerical summary for all continuous variables
summary(cont_vars)
```
### Temperature (`temp`)  

```{r}
hist(bike_2011$temp, main = "Histogram of Normalised Temperature",
     xlab = "temp", breaks = 20)
```

### Feeling Temperature (`atemp`)  

```{r}
hist(bike_2011$atemp, main = "Histogram of Normalised Feeling Temp",
     xlab = "atemp", breaks = 20)
```

### Humidity (`hum`)  

```{r}
hist(bike_2011$hum, main = "Histogram of Normalised Humidity",
     xlab = "hum", breaks = 20)
```

### Wind Speed (`windspeed`)  

```{r}
hist(bike_2011$windspeed, main = "Histogram of Normalised Wind Speed",
     xlab = "windspeed", breaks = 20)
```

### Total Rentals (`cnt`)  

```{r}
hist(bike_2011$cnt, main = "Histogram of Total Rentals",
     xlab = "cnt", breaks = 25)
```

# 2 - Fit linear regression models predicting `cnt` using each of the variables separately.  
## cnt ~ season  
```{r}
model_season <- lm(cnt ~ season, data = bike_2011)
summary(model_season)
```
The model explains about 57% of the variance in total rentals. The coefficients indicate that, compared to Spring, summer has 2108 more rentals, fall has 2797 more rentals, and winter has 1997 more rentals on average. Overall, season is a good predictor of total rentals.  

## cnt ~ mnth  
```{r}
model_mnth <- lm(cnt ~ mnth, data = bike_2011)
summary(model_mnth)
```
The model has a better R2 than season, explaining about 70% of the variance in total rentals. Due to the strong corelation between month and season, the coefficients for month show a clear seasonal pattern. Rentals peak in the summer months (6, 7, 8).  

## cnt ~ holiday  
```{r}
model_holiday <- lm(cnt ~ holiday, data = bike_2011)
summary(model_holiday)
```
The model has a very low R2 of 0.002, indicating that holiday status is not a good predictor of total rentals.  

## cnt ~ weekday  
```{r}
model_weekday <- lm(cnt ~ weekday, data = bike_2011)
summary(model_weekday)
```
This model also has a low R2 of 0.003, suggesting that day of the week is a bad predictor of total rentals.  

## cnt ~ workingday
```{r}
model_workingday <- lm(cnt ~ workingday, data = bike_2011)
summary(model_workingday)
```

This model's R2 is worse. At 0.0004, it indicates that whether it's a working day or not has no predictive power for total rentals.  

## cnt ~ weathersit  
```{r}
model_weathersit <- lm(cnt ~ weathersit, data = bike_2011)
summary(model_weathersit)
```
This model has an R2 of 0.11. The coefficients show that, compared to clear weather, adverse weather conditions (mist/cloudy, light rain/snow, heavy rain/snow) are associated with fewer rentals. However, the model's overall predictive power is still quite low.  

## cnt ~ temp  
```{r}
model_temp <- lm(cnt ~ temp, data = bike_2011)
summary(model_temp)
```
This model has an R2 of 0.59, indicating that temperature is a strong predictor of total rentals. The positive coefficient suggests that higher temperatures are associated with more rentals.  

## cnt ~ atemp  
```{r}
model_atemp <- lm(cnt ~ atemp, data = bike_2011)
summary(model_atemp)
```
This model has an R2 of 0.6, slightly better than temp. The positive coefficient indicates that higher feeling temperatures are associated with more rentals.  

## cnt ~ hum  
```{r}
model_hum <- lm(cnt ~ hum, data = bike_2011)
summary(model_hum)
```

This model has a very low R2, the p-value for hum is also not significant. This suggests that humidity is not a good predictor of total rentals.  

## cnt ~ windspeed  
```{r}
model_windspeed <- lm(cnt ~ windspeed, data = bike_2011)
summary(model_windspeed)
```
This model has a low R2 pf 0.07, indicating that wind speed is a weak predictor of total rentals.  

# 3 - Fit multiple linear regression models with interactions.  
Given that season, mnth, weathersit, temp, atemp are good predictors individually, we will explore interactions between these variables.  
## cnt ~ temp*weathersit  
```{r}
model_temp_weathersit <- lm(cnt ~ temp * weathersit, data = bike_2011)
summary(model_temp_weathersit)
```
This model has an R2 of 0.67, which is an improvement over the temp-only model. The interaction terms are not significant, suggesting that the effect of temperature on rentals does not differ significantly across weather situations. However, the main effect of temp and cloudy weather remain strong and significant.  

## cnt ~ temp*season  
```{r}
model_temp_season <- lm(cnt ~ temp * season, data = bike_2011)
summary(model_temp_season)
```
This model has an R2 of 0.71, which is better than both the temp-only and season-only models. The interaction terms are significant, indicating that the effect of temperature on rentals varies by season.  

## cnt ~ month*weathersit  
```{r}
model_month_weathersit <- lm(cnt ~ mnth * weathersit, data = bike_2011)
summary(model_month_weathersit)
```
This model produces the best R2 so far of 0.814. All of the months are significant predictors of rentals, and some interaction terms are also significant, indicating that the effect of month on rentals varies by weather situation. This makes sense as the impact of weather on rentals is likely to be more pronounced in certain months (e.g cloudy weather may have higher effect in summer months when it's usually hot).  


# 4. Using Backward Selection

```{r}
# we need to drop the empty weathersit level first
bike_2011 <- bike_2011 %>% mutate(weathersit = droplevels(weathersit))

# from part 3, we use the interaction plus all the predicts
full_model <- lm(cnt ~ season + mnth + holiday + weekday + workingday + weathersit + temp + hum + 
windspeed + temp:weathersit + temp:season + 
mnth:weathersit,
data = bike_2011)

summary(full_model)

# using AIC for backward selection
library(MASS)

final_model <- stepAIC(full_model, direction = "backward", trace = TRUE)

summary(final_model)
formula(final_model)
```

- So we first start with the full model, we get AIC of 4590.85. Then we first dropped weekday and because removing it, our model improved to 4584.7, that means weekday added noises not signal. So then we dropped temp:weathesit interaction, it then improved to 4584.0. So then we stopped because removing anything else would increasing the AIC. Hence the final model we keep is cnt ~ season + mnth + holiday + weathersit + temp + hum + windspeed + season:temp + mnth:weathersit. So to my undersanding this part is that AIC balances between model fit and complexity. Basically it only keeps a predictor if it earns its place by improving fit enough to justify the added complexity. And so as we see that weekday and temp:weathersit failed the backwward selection test.

- Another thing, the final model explains 88 percent of variance in daily rental as shown R^2 = 0.88. Temperature is the strongest driver of ridership, however, it varies by season. So as we can see in Fall where rental drops sharply as temperatures cool. Huminity, windspeed, and holidays all negatively impact rentals. Bad weather like Mist/CLoudy, light rain/snow is hurting ridership most during peak summer and fall months when they expect it to be good weather.

# 5. 
```{r}
# preparing data for 2012
bike_2012 <- bike_raw %>% filter(yr == 1) %>% mutate(weathersit = droplevels(weathersit))
```

```{r}
# predict and print out the metrics
preds <- predict(final_model, newdata = bike_2012)

# metrics
rmse <- sqrt(mean((bike_2012$cnt - preds)^2))
r2   <- cor(bike_2012$cnt, preds)^2

rmse
r2
```
-The model performs quite worse comparing to last part. This suggests us that model captured 2011 data pattern well but does not generalize perfectly to 2012. But think about this intuitively, it is likely that ridership grew overall in 2012 as the bike share system matued. So it made sense that a trend the model was not trained to account for.

```{r}
# visualize the plot
ggplot(bike_2012, aes(x = preds, y = cnt)) + geom_point(alpha = 0.4, color = "steelblue") +
geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
labs(title = "Actual vs Predicted - 2012", x = "Predicted", y = "Actual") +
theme_minimal()
```



