---
title: "DATA 5321 - Group Final Project"
author: "Duy Nguyen"
date: "2026-03-04"
output:
  pdf_document:
    latex_engine: xelatex
geometry: margin=0.7in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 60),  # Try 60-70 instead of 40
  size = "small"  # or "footnotesize" for smaller
)

library(tidyverse)
```

```{r}
# Read Data
bike_raw <- read.csv("/Users/dcnguyen060899/Downloads/SU_Winter_2026/DATA_5321/DATA5321_Bike_Share_Daily_Data/data/bike_sharing_daily.csv")

# Re-assign values of variables
bike_raw <- bike_raw %>%
  mutate(
    season     = factor(season,     levels = 1:4,  labels = c("Spring", "Summer", "Fall", "Winter")),
    mnth       = factor(mnth,       levels = 1:12),
    holiday    = factor(holiday,    levels = c(0,1), labels = c("No", "Yes")),
    weekday    = factor(weekday,    levels = 0:6,  labels = c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")),
    workingday = factor(workingday, levels = c(0,1), labels = c("No", "Yes")),
    weathersit = factor(weathersit, levels = 1:4,  labels = c("Clear", "Mist/Cloudy", "Light Rain/Snow", "Heavy Rain/Snow"))
  )

# Filter for 2011
bike_2011 <- bike_raw %>% filter(yr == 0)
```

```{r}
# let's first understand the data dimension, variable type, any missing value, its basic summary statistic
dim(bike_2011)
str(bike_2011)
summary(bike_2011)
colSums(is.na(bike_2011))

```
 - There are 365 rows, so that is basically 365 days in a year in 2011, and 16 columns
 - Factors are correctly assigned to categorical feature like season, mnth, holiday, weekday, workingday, weathersit
 - our outcome variable are cnt, casual, registed and they are integers
 - instant is like ID indexing and yr encode as year like 0 for 20113, 1 for 2012, so these we probably won't include as our predictors.

For summary:
- 90 days each for each seasons so data is balanced
- weathersit:mostly clear with 226, Mist/Claudy (124), very few Light Rain (15), zero Heavy Rain

Good news, no missing data.

However, considering three things:
1. There is no observation heavy rain or snow. Might bias in modelling.
2. hum minimum is 0 seems unsual
3. cnt has a wide range, good to check its distribution in EDA

Now we have a set of continuous and categorical variables.

```{r}
# Continuous variables
bike_2011 %>%
  select(temp, atemp, hum, windspeed, cnt) %>%
  summary()

# Standard deviations
bike_2011 %>%
  select(temp, atemp, hum, windspeed, cnt) %>%
  summarise(across(everything(), sd))

# Categorical variables - counts and proportions
cat_vars <- c("season", "mnth", "holiday", "weekday", "workingday", "weathersit")

for(var in cat_vars) {
  cat("\n---", var, "---\n")
  tbl <- table(bike_2011[[var]])
  print(tbl)
  print(round(prop.table(tbl), 3))
}
```

- Important thing to flag:
1. again hum min is 0, this could be possible outlier
2. again heavy rain or snow have no observation
3. holiday is very imbalanced
4. then also temp and atemp is likely multicollinear (this would violate the “no multicollinearity” (full-rank X) assumption of linear regression)

```{r}
cont_vars <- c("temp", "atemp", "hum", "windspeed", "cnt")

for(var in cont_vars) {
  p <- ggplot(bike_2011, aes(x = .data[[var]])) +
    geom_density(fill = "steelblue", alpha = 0.5) +
    geom_vline(aes(xintercept = mean(.data[[var]])),
               color = "red", linetype = "dashed") +
    labs(title = paste("Density Plot of", var),
         x = var, y = "Density") +
    theme_minimal()
  print(p)
}
```

- temp: two peak, indicating bimodal
- atemp: also bimodal, indicating it correlate with tempt
- hum: left skewed, though look like a small bum at 0 even though stat said otherwise
- windspeed: right skewed
- cnt: also look bimodal, similar mirros the temp because intuitively, code days are low rental and warm days high rentals.

```{r}

for(var in cat_vars) {
  p <- ggplot(bike_2011, aes(x = .data[[var]])) +
    geom_bar(fill = "steelblue", alpha = 0.7) +
    geom_text(stat = "count",
              aes(label = after_stat(count)),
              vjust = -0.5, size = 3) +
    labs(title = paste("Frequency of", var),
         x = var, y = "Count") +
    theme_minimal()
  print(p)
}
```

- season: near perfectly balanced
- mnth: also seem balanced too
- holiday: severely imbalanced
- weekday: perfectly balanced
- workingday: moderately balanced. Though note that workinday and weekday + holiday are logically redundant because working day is determined by weekday and holiday
- Since again, we repeated ourself that heavy rain/snow completely absent so let's drop it for modelling

Key flag to watch out:
- holiday imbalance - unreliable predictor
- workingday is multicollinearity with weekday + holiday

```{r}
# dropping heavy rain/snow
bike_2011 <- bike_2011 %>%
  mutate(weathersit = droplevels(weathersit))

levels(bike_2011$weathersit)
```

Now let's understand the relationship between predictors and cnt?

```{r}
cont_preds <- c("temp", "atemp", "hum", "windspeed")

for(var in cont_preds) {
  p <- ggplot(bike_2011, aes(x = .data[[var]], y = cnt)) +
    geom_point(alpha = 0.4, color = "steelblue") +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(title = paste("cnt vs", var),
         x = var, y = "Daily Rentals (cnt)") +
    theme_minimal()
  print(p)
}

```

- cnt vs temp: there is a strong positive relationship
- cnt vs atemp: nearly identical to temp plot, so this confim multicollinearity
- cnt vs hum: almost a flat line indicating no relationship
- cnt vs windspeed: there is a negative relationship



```{r}
cat_preds <- c("season", "mnth", "holiday", 
               "weekday", "workingday", "weathersit")

for(var in cat_preds) {
  p <- ggplot(bike_2011, aes(x = .data[[var]], y = cnt)) +
    geom_boxplot(fill = "steelblue", alpha = 0.5) +
    labs(title = paste("cnt by", var),
         x = var, y = "Daily Rentals (cnt)") +
    theme_minimal()
  print(p)
}
```

- There is a clear differences across seasons. Because of this differences, there will be variations model can expoloit therefore this going to be a strong predictor.
- cnt by mnth: There is an inverted U-shape. I would say that mnth captures the seasonal trend very granularly

- cnt by holiday: there is a slight differnce as holidays have slightly lower median rentals. the differnce is not significant. I would say this is a week predictor.
- cnt by weekday: boxes are very similar across all days. Hence, there is a little effect on rental. probably week predictor as well
- cnt by workingday: identical distributions. probably week predictor as well.
- weatherit: after we dropped the no observation for heavy rain/snow, there is a clear downward staircase pattern. That means as weather worsen so rental drop

Note that Light Rain/Snow box is very wide relative to its size, it would be unstable estimates due to only 15 observations

### Summary:
```{r, echo=FALSE}
cat("
| Predictor      | Type         | Relationship with cnt                          |
|----------------|--------------|-----------------------------------------------|
| temp / atemp   | Continuous   | Strong positive                               |
| season         | Categorical  | Strong — clear group differences              |
| mnth           | Categorical  | Strong — inverted U shape                     |
| weathersit     | Categorical  | Strong — staircase drop                       |
| windspeed      | Continuous   | Moderate negative                             |
| hum            | Continuous   | Weak/flat                                     |
| holiday        | Categorical  | Weak — overlapping boxes                      |
| weekday        | Categorical  | Weak — nearly identical boxes                 |
| workingday     | Categorical  | Weak — nearly identical boxes                 |
")
```



